Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-07-18 paper.tex
Fix sign error in optical depth eqn
Commit a bunch of small fixes from the weekend

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{deat2013-leakage, pere2016-asl, radi2016-dynamical}
\todo{cite others}
capture the qualitative effects of neutrinos on the matter, but
provide extremely limited information about the neutrino field itself.
\todo{discuss mc more clearly}
Monte Carlo methods \citep{abdi2012-monte_carlo, rich2015-monte_carlo}
are an excellent tool require large computational resources because of
their global treatment of the problem.

The state of the art today couples radiation and matter with a truncated
moment formalism
\citep{shib2011-truncated_moment, fouc2015-m1_nsbh, fouc2016-m1_nsns,
  just2015-m1_code},
\todo{cite others}
evolving the zeroth- and first-angular moments of the distribution function
(the energy density and flux respectively) representing the zeroth-energy moment
of the radiation field (the total energy).
This method, commonly called M1 transport, was formulated by
\cite{thor1981-truncated_moment, shib2011-truncated_moment}
and has been applied for example in \cite{ramp2002-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}.
Recently, \cite{fouc2016-m1_evolve_n} have expanded their M1 transport code
to also evolve the number densities of neutrinos,
providing total and average energies throughout the simulated volume.
Even so, with M1 transport codes only evolving two angular moments
and one or two energy moments,
we can extract only limited angular and spectral information
from the neutrino fields.

@@

\subsection{The Rendering Equation}
\label{ssec:rendering_eqn}
We can integrate Eqn.~\ref{eqn:boltzmann_split} directly, 
with our solution split into a boundary, absorption/emission, and
a scattering term, $f=f_{\rm bdry}+f_{\rm AE}+f_{\rm SE}$:
\begin{align}
  \label{eqn:rendering_fbdry}
  f_{\rm bdry}(\lambda,\lambda_e)
  &= f(\lambda_e) e^{-\tau(\lambda,\lambda_e)}, \\
  \label{eqn:rendering_fae}
  f_{\rm AE}(\lambda,\lambda_e)
  &= \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{K}_{\rm AE}(\lambda')f^{\rm eq}(\lambda'), \\
  \label{eqn:rendering_fse}
  f_{\rm SE}(\lambda,\lambda_e)
  &= \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{K}_{\rm SE}(\lambda')\Phi(\lambda'),
\end{align}
where the optical depth is defined,
\begin{equation}
  \label{eqn:optical_depth}
  \tau(\lambda,\lambda') \equiv -\int_{\lambda'}^\lambda
  d\lambda'' \, \mathscr{K}(\lambda''),
\end{equation}
and the parameterization conventions are depicted in
Fig.~\ref{fig:affine_param}.
Note that Eqn.~\ref{eqn:optical_depth} employs the total absorption
plus scattering opacity, so that the optical depth attenuating the
integrands of Eqns.~\ref{eqn:rendering_fae} and \ref{eqn:rendering_fse}
is the total optical depth.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figure-mfps-50km}
  \caption{Mean free paths
    representative of the collapse profile at
    50~km ($\varrho=10^{11}\,{\rm g}\,{\rm cm}^{-3}$,
    $T=3.7$~MeV, $Y_e=0.12$).
    In the fluid rest frame these are given by
    $(\chi^*_a)^{-1}$ for the absorption
    and $(\chi_s)^{-1}$ for elastic scattering interactions.
    Computed using \lstinline{NuLib}, as in the ray tracing code.
  }
  \label{fig:mfps-50km}
\end{figure}

@@

\subsection{Hypermassive Neutron Star Remnant:
  Testing Everything}
\label{ssec:test_disk_comparison}
The merger of two neutron stars by gravitational wave emission produces a
postmerger configuration composed of a single neutron star surrounded by a disk.
Because of the strong differential rotation of the remnant it temporarily avoids
collapse to a black hole, even if its mass exceeds the threshold of dynamical
instability for a rigidly rotating neutron star \cite{duez2009-review}.
These objects are called hypermassive neutron stars.

Such a configuration was modeled by \cite{fouc2016-m1_nsns} by
evolving the final inspiral and merger of two identical neutron stars of
isolated gravitational mass 1.2~$M_{\odot}$. The postmerger configuration
was evolved using an M1 transport scheme for the neutrinos, evolving the
energy density, number density, and energy flux \cite{fouc2016-m1_evolve_n}
for $\sim10\,{\rm ms}$ following merger, defined as the time of the peak of the
gravitational wave strain.

We used a single time snapshot of the volume data from that configuration
at $t=11\,{\rm ms}$ after merger (see
Figs.~\ref{fig:nsns_rho_merid}--\ref{fig:nsns_temp_equat})
sampling the neutrino fields with ray tracing to compare to some
measures of the field computed from the M1 transport simulation.

@@

The merger models are described in \cite{fouc2015-m1_nsbh, fouc2016-m1_nsns}.
Something like this is presented in \cite[Figs.~10-11]{pere2014-nu_wind};
compare.

@@

In practice we use two possible fiducial observers to define the momentum
decomposition via Eqn.~\ref{eqn:def_momentum}: the Eulerian,
or normal observer $n_\mu=-\alpha \,\partial_\mu t$,
and the fluid, or comoving observer
$u^\mu=Wn^\mu+v_{\rm E}^\mu$.
Here $t$ is coordinate time and $\alpha$ is lapse
in the standard 3+1 decomposition of the metric
\begin{equation}
  \label{eqn:adm_metric}
  \psi_{\mu\gamma} \rightarrow
  \left(
  \begin{matrix}
    -\alpha^2 + \beta^i \beta_i  & \beta_i \\
    \beta_j                      & g_{ij}
  \end{matrix}
  \right).
\end{equation}
We have also introduced
the fluid Lorentz factor $W=-n_\mu u^\mu=\alpha u^t$,
its Eulerian velocity $v_{\rm E}^\mu=g^\mu_\lambda u^\lambda$
(distinct from its coordinate velocity $v^\mu=u^\mu/u^t$),
and the projection tensor orthogonal to the normal observer
$g^\mu_\lambda=\psi^\mu_\lambda+n^\mu n_\lambda$,
whose spatial part is the 3-metric, $g_{ij}$.

>>>>

This is a challenging task in the environment of a merger,
which generally lacks any spatial symmetries, so that fully general
seven-dimensional solutions to the Boltzmann Equation are not feasible.
Leakage approximations
\citep{deat2013-leakage, pere2016-asl, radi2016-dynamical}
\todo{cite others}
capture the qualitative effects of neutrinos on the matter, but
provide extremely limited information about the neutrino field itself.
\todo{discuss mc more clearly}
Monte Carlo methods \citep{abdi2012-monte_carlo, rich2015-monte_carlo}
are an excellent tool but require large computational resources because of
their global treatment of the problem.

The state of the art today couples radiation and matter using a truncated
moment formalism
\citep{shib2011-truncated_moment, fouc2015-m1_nsbh, fouc2016-m1_nsns,
  just2015-m1_code},
\todo{cite others}
evolving the zeroth- and first-angular moments of the energy density,
that is the total energy density and flux.
This method, commonly called M1 transport, was formulated by
\cite{thor1981-truncated_moment, shib2011-truncated_moment}
and has been applied for example in \cite{ramp2002-truncated_moment,
  card2013-truncated_moment, fouc2015-m1_nsbh, ocon2015-gr1d_with_nu}.
Recently, \cite{fouc2016-m1_evolve_n} have expanded their M1 transport code
to also evolve the zeroth-angular moment of the number density,
providing total energy densities and average energies throughout the
simulated volume.
Even so, with M1 transport codes only evolving two angular moments
and one or two energy moments,
they can extract only limited angular and spectral information
from the neutrino fields.

@@

\subsection{The Rendering Equation}
\label{ssec:rendering_eqn}
We can integrate Eqn.~\ref{eqn:boltzmann_split} directly, 
with our solution split into a boundary, absorption/emission, and
a scattering term, $f=f_{\rm bdry}+f_{\rm AE}+f_{\rm SE}$:
\begin{align}
  \label{eqn:rendering_fbdry}
  f_{\rm bdry}(\lambda,\lambda_e)
  &= f(\lambda_e) e^{-\tau(\lambda,\lambda_e)}, \\
  \label{eqn:rendering_fae}
  f_{\rm AE}(\lambda,\lambda_e)
  &= \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{K}_{\rm AE}(\lambda')f^{\rm eq}(\lambda'), \\
  \label{eqn:rendering_fse}
  f_{\rm SE}(\lambda,\lambda_e)
  &= \int_{\lambda_e}^\lambda d\lambda'\,e^{-\tau(\lambda,\lambda')}
  \mathscr{K}_{\rm SE}(\lambda')\Phi(\lambda'),
\end{align}
where the optical depth is defined,
\begin{equation}
  \label{eqn:optical_depth}
  \tau(\lambda,\lambda') \equiv \int_{\lambda'}^\lambda
  d\lambda'' \, \mathscr{K}(\lambda''),
\end{equation}
and the parameterization conventions are depicted in
Fig.~\ref{fig:affine_param}.
Note that Eqn.~\ref{eqn:optical_depth} employs the total absorption
plus scattering opacity, so that the optical depth attenuating the
integrands of Eqns.~\ref{eqn:rendering_fae} and \ref{eqn:rendering_fse}
is the total optical depth.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figure-mfps-50km}
  \caption{Mean free paths
    representative of the collapse profile at
    50~km ($\varrho=10^{11}\,{\rm g}\,{\rm cm}^{-3}$,
    $T=3.7$~MeV, $Y_e=0.12$).
    In the fluid rest frame these are given by
    $(\chi^*_a)^{-1}$ for the absorption
    and $(\chi_s)^{-1}$ for elastic scattering interactions.
    Note that elastic scattering opacities are identical across species
    below energies at which weak magnetism plays a role.
    In this data they are exactly identical
    because we have turned off weak magnetism in our opacity calculations
    in order to compare our results with the historical literature.
    Computed using \lstinline{NuLib}.
  }
  \label{fig:mfps-50km}
\end{figure}

@@

\subsection{Hypermassive Neutron Star Remnant:
  Testing Everything}
\label{ssec:test_disk_comparison}
The merger of two neutron stars by gravitational wave emission produces a
postmerger configuration composed of a single neutron star surrounded by a disk.
Because of the strong differential rotation of the remnant it temporarily avoids
collapse to a black hole, even if its mass exceeds the threshold of dynamical
instability for a rigidly rotating neutron star \cite{duez2009-review}.
These objects, called hypermassive neutron stars,
may avoid collapse for $X\textrm{--}Y$~s,
\todo{cite $X$, $Y$}
depending upon a number of physical factors including thermal pressure,
magnetic fields, and the microphysics of the fluid.

Such a configuration was modeled in \cite{fouc2016-m1_nsns} by
evolving the final inspiral and merger of two identical neutron stars of
isolated gravitational mass 1.2~$M_{\odot}$. The postmerger configuration
was simulated using an M1 transport scheme for the neutrinos, evolving the
energy density, number density, and energy flux \cite{fouc2016-m1_evolve_n}
for $\sim10\,{\rm ms}$ following merger.
The fluid was modeled using the LS220 equation of state \cite{latt1991-nuc_eos}.

We use a single time snapshot of the volume data from that configuration
at $t=11\,{\rm ms}$ after merger (see
Figs.~\ref{fig:nsns_rho_merid}--\ref{fig:nsns_temp_equat})
sampling the neutrino fields with ray tracing to compare to some
measures of the field computed from the M1 transport simulation.
\todo{here focus on plots as functions of $\theta$}

@@

The merger models are described in \cite{fouc2015-m1_nsbh, fouc2016-m1_nsns}.
Something like this is presented in \cite[Figs.~10-11]{pere2014-nu_wind};
compare.
\todo{here focus on plots of $J(\cos A)$, $H^r(\cos A)$, and $J(\varepsilon)$}

@@

We use two possible fiducial observers to define the momentum
decomposition via Eqn.~\ref{eqn:def_momentum_2}: the Eulerian,
or normal observer $n_\mu=-\alpha \,\partial_\mu t$,
and the fluid, or comoving observer
$u^\mu=Wn^\mu+v_{\rm E}^\mu$.
Here $t$ is coordinate time and $\alpha$ is lapse
in the standard 3+1 decomposition of the metric
\begin{equation}
  \label{eqn:adm_metric}
  \psi_{\mu\gamma} \rightarrow
  \left(
  \begin{matrix}
    -\alpha^2 + \beta^i \beta_i  & \beta_i \\
    \beta_j                      & g_{ij}
  \end{matrix}
  \right).
\end{equation}
We have also introduced
the fluid Lorentz factor $W=-n_\mu u^\mu=\alpha u^t$,
its Eulerian velocity $v_{\rm E}^\mu=g^\mu_\lambda u^\lambda$
(distinct from its coordinate velocity $v^\mu=u^\mu/u^t$),
and the projection tensor orthogonal to the normal observer
$g^\mu_\lambda=\psi^\mu_\lambda+n^\mu n_\lambda$,
whose spatial part is the 3-metric, $g_{ij}$.
